---
title: "Trade data analysis"
author: "TH"
format: html
execute-dir: file
execute:
  echo: true
  warning: false
  message: false
  output: false
editor: visual
---

# IAS presence in e-commerce establishments

### Packages

```{r}
library(ggplot2)
library(dplyr)
library(stringr)
library(tidyr)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(cowplot)
library(patchwork)
```

## Define functions

```{r}
clean_species <- function(x){
  x %>% 
    # Remove question marks and asterisks
    str_remove("[?]+") %>% 
    str_remove("[*]+") %>% 
    
    # Remove all brackets with synonym names
    str_remove("[(][^()]+[)]") %>%
    
    # Remove excessive spaces
    str_squish() 
}
```

## Loading data

```{r}
species.2021 <- read.csv("data/species_list_2021.csv")
species.2024 <- read.csv("data/species_list_2024.csv")

# Trading records
unique.records.2021 <- read.csv("data/unique_records_2021.csv") %>% 
  rename_with(tolower) %>% 
  select(species, platform, kingdom, country) %>% 
  mutate(year = "2021")

unique.records.2024 <- read.csv("data/unique_records_2024.csv") %>% 
  rename_with(tolower) %>% 
  select(species, platform, kingdom, country) %>% 
  mutate(year = "2024")

# surveyed species in 2021
surveyed.in.2021 <- unique.records.2021 %>% 
  pull("species") %>% 
  clean_species() %>% 
  unique()

# traded species in 2021
traded.in.2021 <- read.csv("data/results_2021.csv") %>% 
  pull("Species") %>%
  clean_species() %>% 
  unique()

# surveyed species in 2024
surveyed.in.2024 <- unique.records.2024 %>% 
  pull("species") %>% 
  clean_species() %>% 
  unique()

# traded species in 2024
traded.in.2024 <- read.csv("data/results_2024.csv") %>% 
  filter(Total.number.establishments.found.in > 0) %>% 
  pull("Species") %>% 
  clean_species() %>% 
  unique()
```

### Combine data

```{r}
#select and rename columns of 2021 and 2024 data
species.2021.sel <- species.2021 %>%
  rename_with(tolower) %>% 
  select(
    species = scientific.name,
    kingdom,
    status.2021 = union.concern.status,
    sector.2021 = sector
  ) %>% 
  mutate(species_clean = clean_species(species))

species.2024.sel <- species.2024 %>%
  rename_with(tolower) %>% 
  select(
    species,
    kingdom,
    status.2024 = status,
    sector.2024 = sector,
  ) %>% 
  mutate(species_clean = clean_species(species))

#to check alignment of species list
## This works only in case the list is ordered!
# identical(species.2021.sel$species, species.2024.sel$species) #returns NO
setdiff(species.2021.sel$species, species.2024.sel$species) #highlights 3 species that were proposed in 2021 but not included in 2024
setdiff(species.2024.sel$species, species.2021.sel$species) #highlights the 22 species that were added as "proposed" to the 2024 list


#combine in one df
species.combined <- species.2024.sel %>%
  full_join(species.2021.sel) %>%
  mutate(
    found.in.trade.2021 = if_else(species_clean %in% traded.in.2021, "Yes","No"),
    found.in.trade.2024 = if_else(species_clean %in% traded.in.2024, "Yes", "No"),
    status.2024 = case_when(
      str_detect(status.2024, "^Union concern") ~ "Union concern",
      status.2024 == "Proposed" ~ "Proposed"
    ),
  )

```

## Classify species into Entry, Persistent, and Exit

The classification takes into account the species' status in 2021 and 2024 as well as whether it was considered for analysis in 2021.

This results in the following 7 classifications based on the decision tree: Included in 2021? (Yes/No) - Status 2021? (NA/Proposed/Union concern) - Found in trade in 2021? (NA/No/Yes) - Status 2024? (Proposed/Union concern) - Found in trade in 2024? (No/Yes) (**Please note:** 4 species are classified as "Union concern delayed to 2024/2027" but were considered as listed on the Union concern list for the analysis; all were persistent-proposed)

**Entry - New:** Species was not included in 2021 survey but was listed included as a proposed IAS in 2024 and found in trade

**Entry - Proposed:** Species was included as a proposed IAS in 2021 survey. It was not identified in any e-commerce establishments in 2021 but was documented in 2024 as a listed IAS present in at least one e-commerce establishment.

**Entry - Listed:** Species was already included as a listed IAS in 2021. While it was not documented in any e-commerce establishment that year, it was documented in at least one establishment in 2024.

**Persistent - Proposed:** Species was identified on at least one establishment in both years. While it was included as a proposed IAS in 2021 it was categorized as a listed IAS in 2024.

**Persistent - Listed:** Species was identified as listed IAS in at least one establishment in 2021 and 2024.

**Exit - Proposed:** Species was identified as a proposed IAS in at least one establishment in 2021 but categorized as a listed IAS in the 2024 survey it was not documented in any establishment.

**Exit - Listed:** Species was categorized as listed IAS in both surveys but was identified in at least one establishment only in 2021.

```{r}
species.classified <- species.combined %>%
  mutate(
    category = case_when(
      # Species was traded in both years
      found.in.trade.2021 == "Yes" & found.in.trade.2024 == "Yes" ~ if_else(
        status.2021 == "Proposed", "Persistent - Proposed","Persistent - Listed"
      ),
      
      # Species was traded just in 2024
      found.in.trade.2021 == "No" & found.in.trade.2024 == "Yes" ~ case_when(
        # Species was not surveyed in 2021
        is.na(status.2021) ~ "Entry - New",
        
        # Species was surveyed in 2021 as proposed
        status.2021 == "Proposed" ~ "Entry - Proposed",
        
        # Species was surveyed in 2021 as union concern
        status.2021 == "Union concern" ~ "Entry - Listed",
      ), 
      
      # Species was traded just in 2021
      found.in.trade.2021 == "Yes" & found.in.trade.2024 == "No" ~ case_when(
        # Species was surveyed as proposed in 2021
        status.2021 == "Proposed" ~ if_else(is.na(status.2024), "Exit - New", "Exit - Proposed"),
        
        # Species was surveyed as union concern in 2021
        status.2021 == "Union concern" ~ "Exit - Listed"
      ), 
  
      TRUE ~ "No trade")) %>% 
  # Convert category to factor with specified order
  mutate(across(category, ~factor(.x, levels = c(
    "Entry - New", "Entry - Proposed", "Entry - Listed",
    "Persistent - Proposed", "Persistent - Listed",
    "Exit - Proposed", "Exit - Listed", "Exit - New", "No trade"
  ))))
```

### Visualization of species classification

```{r}

#create counts to include all categories in graph
plot_data <- species.classified %>%
  count(category, kingdom) %>% 
  complete(category, kingdom, fill = list(n = 0))

#defining base colors for categories
base_colors <- c(
  "Entry - New" = "#1b9e77",
  "Entry - Proposed" = "#66c2a5",
  "Entry - Listed" = "#a6dba0",
  "Persistent - Proposed" = "#ffd92f",
  "Persistent - Listed" = "#e6ab02",
  "Exit - Proposed" = "#377eb8",
  "Exit - Listed" = "#74add1",
  "Exit - New" = "red",
  "No trade" = "grey"
)

#grah the data, indicating the differences between plants & animals with differences in alpha; adding count numbers to graph
ggplot(plot_data, aes(x = category, y = n, fill = category)) +
  geom_col(position = "stack") +
  geom_text(
    aes(label = n),
    position = position_stack(vjust = 0.5),
    size = 3
  ) +
  coord_flip() +
  scale_fill_manual(values = base_colors, drop = FALSE,
                    name = "Classification") +
  theme_minimal() +
  labs(
    title = "Species Classification",
    x = "Category",
    y = "Count"
  ) +
  facet_wrap(.~kingdom)

```

## Visalization of the temporal development

```{r}
#defining colors for displaying Plants vs Animals
colors <- c(
  "Plant" = "#1b9e77",
  "Animal" = "#377eb8"
)

#adjusting the data to explore the development of species presence (yes/no) in trade
plot_species <- bind_rows(unique.records.2021, unique.records.2024) %>%
  filter(!str_detect(species, "sp[p]*[.]$"))

plot_records <- plot_species %>% 
  count(country, year, platform, kingdom) %>%
  mutate(across(kingdom,~factor(.x, levels = c("Plant", "Animal"))))

```

### Creating the underlying map

```{r}

europe <- ne_countries(scale = "medium", continent = "Europe", returnclass = "sf")

unique(europe$name)


# Crop to mainland Europe
europe_cropped <- st_crop(
  europe,
  xmin = -10, xmax = 35,
  ymin = 34, ymax = 72
)


countries <- c("Germany", "Spain", "Czechia", "France", "Italy", "Netherlands")

map_data <- europe_cropped %>%
  filter(name %in% countries)

centroids <- st_centroid(map_data)

centroids_df <- as.data.frame(st_coordinates(centroids))
centroids_df$Country <- map_data$name

#since we have "International data, we will add a manual location for it on the map
centroids_df <- rbind(
  centroids_df,
  data.frame(
    X = 25,   # longitude (adjust visually)
    Y = 45,   # latitude (adjust visually)
    Country = "International"
  )
)


```

### Creating functions

i\) a function to create individual stacked bar plots per country. The plots indicate the composition of animal and plant species found in Online retailers (left bar) and Peer-to-peer trade platforms (right) in 2021 and 2024

ii\) a function to display the individual bar plots on the underlying map

```{r}
make_barplot <- function(df, show_legend = FALSE) {
   df <- df %>%
    tidyr::complete(
      Platform,
      year,
      Kingdom = c("Plant", "Animal"),
      fill = list(n = 0)
    )
  ggplot(df, aes(x = Platform, y = n, fill = Kingdom)) +
    geom_col(position = "fill") +
    facet_wrap(~year) +
    scale_fill_manual(values = colors, drop = FALSE) +
     labs(
      x = "Platform",
      fill = "Kingdom"
    ) +
    theme_void()+
    theme(
      legend.position = if (show_legend) "bottom" else "none"
    )
}



make_map_plot <- function(summary_data, title_text) {

  base_map <- ggplot() +
    geom_sf(data = map_data, fill = "grey90", color = "black") +
    theme_void() +
    labs(title = title_text)

  for (i in 1:nrow(centroids_df)) {

    country_name <- centroids_df$Country[i]

    p <- make_barplot(filter(summary_data, Country == country_name))

    # Default offsets
x_offset <- -2.5
x_offset_max <- 2.5
y_offset <- -2.5
y_offset_max <- 2.5

# Netherlands: move left
if (country_name == "Netherlands") {
  x_offset <- -6
  x_offset_max <- -1
  y_offset <- -2
  y_offset_max <- 3
}

# Czechia: move right and up
if (country_name == "Czechia") {
  x_offset <- 1
  x_offset_max <- 6
  y_offset <- 0
  y_offset_max <- 5
}

base_map <- base_map +
  annotation_custom(
    grob = ggplotGrob(p),
    xmin = centroids_df$X[i] + x_offset,
    xmax = centroids_df$X[i] + x_offset_max,
    ymin = centroids_df$Y[i] + y_offset,
    ymax = centroids_df$Y[i] + y_offset_max
  )
  }
base_map <- base_map +
  annotate(
    "text",
    x = 25,
    y = 42,
    label = "International   kf",   #would not print the whole of International otherwise
    size = 3
  )
  return(base_map)
}
```

### Mapping the results

```{r}
map_species <- make_map_plot(
  species.summary,
  "Species Composition in e-commerce"
)

#creating a legend
legend_source <- make_barplot(
  species.summary %>% filter(Country == "Germany"),  # small subset is enough
  show_legend = TRUE
)
legend_only <- get_legend(legend_source)

#adding the legend to the figure
map_bar_species <- plot_grid(
  map_species,
  legend_only,
  ncol = 1,
  rel_heights = c(1, 0.1)
)
```

### Creating a heat map of change

The heat map will display the changes in plant and animal species composition per platform type and country. Specifically, the heat map shows the changes in plant proportions with positive values indicating more plants and negative values indicating more animals.

```{r}
make_heatmap <- function(summary_data, title_text) {

  heat_data <- summary_data %>%
    group_by(Country, year, Platform) %>%
    mutate(prop = n / sum(n)) %>%
    ungroup() %>%
    filter(Kingdom == "Plant") %>%   # adjust if your levels are "Plants"
    select(Country, Platform, year, prop) %>%
    tidyr::pivot_wider(
      names_from = year,
      values_from = prop
    ) %>%
    mutate(change = `2024` - `2021`)

  ggplot(heat_data, aes(x = Platform, y = Country, fill = change)) +
    geom_tile(color = "white") +
    scale_fill_gradient2(
      low = "#377eb8",   # more animals
      mid = "white",
      high = "#1b9e77",  # more plants
      midpoint = 0
    ) +
    labs(
      title = title_text,
      x = "Platform",
      y = "Country",
      fill = "Δ Plant proportion"
    ) +
    theme_minimal()+
  theme(legend.position = "right")
}

heat_species <- make_heatmap(
  species.summary,
  "Change in Plant Proportion (Unique Species) 2021-2024"
)


```
