---
title: "Trade data analysis"
author: "TH"
format: html
execute:
  echo: true
  warning: false
  message: false
  output: false
editor: visual
---

# IAS presence in e-commerce establishments

### Packages

```{r}
library(ggplot2)
library(dplyr)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(cowplot)
library(patchwork)
```

### Set working space

```{r}
setwd("C:/Users/henke/Desktop/library/Projects/trade_data")
```

## Loading data

```{r}
species.2021 <- read.csv("species_list_2021.csv")
species.2024 <- read.csv("species_list_2024.csv")

results.2021 <- read.csv("results_2021.csv")
results.2024 <- read.csv("results_2024.csv")

unique.records.2021 <- read.csv("unique_records_2021.csv")
unique.records.2024 <- read.csv("unique_records_2024.csv")
```

### Combine data

```{r}
#select and rename columns of 2021 and 2024 data
species.2021.sel <- species.2021 %>%
  rename_with(tolower) %>% 
  select(
    species = scientific.name,
    status.2021 = union.concern.status,
    sector.2021 = sector
  )

species.2024.sel <- species.2024 %>%
  rename_with(tolower) %>% 
  select(
    species,
    status.2024 = status,
    sector.2024 = Sector,
    survey.list.2021 = trade.2021.survey.species.list,
    found.in.trade.2021,
    survey.list.2024 = trade.survey.list.2024
  )

#to check alignment of species list
## This works only in case the list is ordered!
# identical(species.2021.sel$species, species.2024.sel$species) #returns NO
setdiff(species.2021.sel$species, species.2024.sel$species) #highlights 3 species that were proposed in 2021 but not included in 2024
setdiff(species.2024.sel$species, species.2021.sel$species) #highlights the 22 species that were added as "proposed" to the 2024 list


#combine in one df
species.combined <- species.2024.sel %>%
  left_join(species.2021.sel, by = "species")

species.combined <- species.combined %>%
  mutate(
    found.in.trade.2021 = case_when(
      Found.in.trade.2021 == "Found in trade" ~ "Yes",
      Found.in.trade.2021 == "Not found in trade" ~ "No",
      # Found.in.trade.2021 == "" ~ NA_character_,
      TRUE ~ NA_character_
    ),
    Status.2024 = case_when(
      str_detect(Status.2024, "^Union concern") ~ "Union concern",
      Status.2024 == "Proposed" ~ "Proposed"
    )
  )


#add which species were found in 2024 in trade (yes/no)
found.in.trade.2021 <- unique(unique.records.2021$Species.found.as.listed.in.species.list)
found.in.trade.2024 <- unique(unique.records.2024$Species.found.as.listed.in.species.list)

#add column indicating whether species was found in trade in 2024
species.combined <- species.combined %>%
  mutate(
    Found.in.trade.2024 = if_else(species %in% found.in.trade.2024, "Yes", "No")
  )
```

## Classify species into Entry, Persistent, and Exit

The classification takes into account the species' status in 2021 and 2024 as well as whether it was considered for analysis in 2021.

This results in the following 7 classifications based on the decision tree: Included in 2021? (Yes/No) - Status 2021? (NA/Proposed/Union concern) - Found in trade in 2021? (NA/No/Yes) - Status 2024? (Proposed/Union concern) - Found in trade in 2024? (No/Yes) (**Please note:** 4 species are classified as "Union concern delayed to 2024/2027" but were considered as listed on the Union concern list for the analysis; all were persistent-proposed)

**Entry - New:** Species was not included in 2021 survey but was listed included as a proposed IAS in 2024 and found in trade

**Entry - Proposed:** Species was included as a proposed IAS in 2021 survey. It was not identified in any e-commerce establishments in 2021 but was documented in 2024 as a listed IAS present in at least one e-commerce establishment.

**Entry - Listed:** Species was already included as a listed IAS in 2021. While it was not documented in any e-commerce establishment that year, it was documented in at least one establishment in 2024.

**Persistent - Proposed:** Species was identified on at least one establishment in both years. While it was included as a proposed IAS in 2021 it was categorized as a listed IAS in 2024.

**Persistent - Listed:** Species was identified as listed IAS in at least one establishment in 2021 and 2024.

**Exit - Proposed:** Species was identified as a proposed IAS in at least one establishment in 2021 but categorized as a listed IAS in the 2024 survey it was not documented in any establishment.

**Exit - Listed:** Species was categorized as listed IAS in both surveys but was identified in at least one establishment only in 2021.

```{r}
species.classified <- species.combined %>%
  mutate(
    Category = case_when(
      # ---- ENTRY ----
      survey.list.2021 == "Not considered" & is.na(Status.2021) & is.na(Found.in.trade.2021) & Status.2024 == "Proposed" & Found.in.trade.2024 == "Yes"                    ~ "Entry - New",
      survey.list.2021 == "Yes" & Status.2021 == "Proposed" & Found.in.trade.2021 == "No" & Status.2024 == "Union concern" & Found.in.trade.2024 == "Yes"                    ~ "Entry - Proposed",
        survey.list.2021 == "Yes" & Status.2021 == "Union concern" & Found.in.trade.2021 == "No" & Status.2024 == "Union concern" & Found.in.trade.2024 == "Yes"                    ~ "Entry - Listed",
      # ---- PERSISTENT ----
      survey.list.2021 == "Yes" & Status.2021 == "Proposed" & Found.in.trade.2021 == "Yes" & Status.2024 == "Union concern" & Found.in.trade.2024 == "Yes"                    ~ "Persistent - Proposed",
      survey.list.2021 == "Yes" & Status.2021 == "Union concern" & Found.in.trade.2021 == "Yes" & Status.2024 == "Union concern" & Found.in.trade.2024 == "Yes"                    ~ "Persistent - Listed",
      # ---- EXIT ----
      survey.list.2021 == "Yes" & Status.2021 == "Proposed" & Found.in.trade.2021 == "Yes" & Status.2024 == "Union concern" & Found.in.trade.2024 == "No"                    ~ "Exit - Proposed",
      survey.list.2021 == "Yes" & Status.2021 == "Union concern" & Found.in.trade.2021 == "Yes" & Status.2024 == "Union concern" & Found.in.trade.2024 == "No"                    ~ "Exit - Listed",
      # ---- NA ----
      survey.list.2021 == "Not considered" & is.na(Status.2021) & is.na(Found.in.trade.2021) & Status.2024 == "Proposed" & Found.in.trade.2024 == "No"                    ~ "No trade",
      survey.list.2021 == "Yes" & Status.2021 == "Proposed" & Found.in.trade.2021 == "No" & Status.2024 == "Union concern" & Found.in.trade.2024 == "No"                    ~ "No trade",
      survey.list.2021 == "Yes" & Status.2021 == "Union concern" & Found.in.trade.2021 == "No" & Status.2024 == "Union concern" & Found.in.trade.2024 == "No"                    ~ "No trade",
      TRUE ~ NA_character_
    )
  ) %>%
  select(species, Kingdom.2024, Category)

table(species.classified$Category, useNA = "ifany")
```

### Visualization of species classification

```{r}
#set category order
species.classified$Category <- factor(
  species.classified$Category,
  levels = c(
    "Entry - New", "Entry - Proposed", "Entry - Listed",
    "Persistent - Proposed", "Persistent - Listed",
    "Exit - Proposed", "Exit - Listed", "No trade"
  )
)

#Create stacked bar graph
species.classified$Category <- factor(
  species.classified$Category,
  levels = c(
    "Entry - New", "Entry - Proposed", "Entry - Listed",
    "Persistent - Proposed", "Persistent - Listed",
    "Exit - Proposed", "Exit - Listed", "No trade"
  )
)

#create counts to include all categories in graph
plot_data <- species.classified %>%
  count(Category, Kingdom.2024) %>%
  tidyr::complete(Category, Kingdom.2024, fill = list(n = 0)) #creates counts

#defining base colors for categories
base_colors <- c(
  "Entry - New" = "#1b9e77",
  "Entry - Proposed" = "#66c2a5",
  "Entry - Listed" = "#a6dba0",
  "Persistent - Proposed" = "#ffd92f",
  "Persistent - Listed" = "#e6ab02",
  "Exit - Proposed" = "#377eb8",
  "Exit - Listed" = "#74add1",
  "No trade" = "grey"
)

#grah the data, indicating the differences between plants & animals with differences in alpha; adding count numbers to graph
ggplot(plot_data, aes(x = Category, y = n, fill = Category, alpha = Kingdom.2024)) +
  geom_col(position = "stack") +
  geom_text(
    aes(label = n),
    position = position_stack(vjust = 0.5),
    size = 3
  ) +
  coord_flip() +
  scale_fill_manual(values = base_colors, drop = FALSE,
                    name = "Classification") +
  scale_alpha_manual(values = c("Plants" = 1.2, "Animals" = 0.5),
                     name = "") +
  theme_classic() +
  labs(
    title = "Species Classification",
    x = "Category",
    y = "Count"
  )



```

## Visalization of the temporal development

### Preparing data

```{r}
#checking consistency and spelling species names and updating if any errors found
#unique(unique.records.2021$Species) #no need for updates
#unique(unique.records.2024$Species)

unique.records.2024 <- unique.records.2024 %>%
  mutate(
    Species = case_when(
      Species == "Ameiurus melas??" ~ "Ameiurus melas",
      Species == "Cabomba caroliniana??" ~ "Cabomba caroliniana",
      Species == "Cherax spp.??" ~ "Cherax spp.",
      Species == "Cortaderia selloana??" ~ "Cortaderia selloana",
      Species == "Elodea nuttallii??" ~ "Elodea nuttallii",
      Species == "Gambusa spp.??" ~ "Gambusia spp.",
      Species == "Gambusia affinis/holbrooki" ~ "Gambusia spp.",
      Species == "Lampropeltis getula??" ~ "Lampropeltis getula",
      Species == "Myriophyllum aquaticum??" ~ "Myriophyllum aquaticum",
      Species == "Myriophyllum heterophyllum??" ~ "Myriophyllum heterophyllum",
      Species == "Pennisetum setaceum??" ~ "Pennisetum setaceum",
      Species == "Procambarus fallax f. virginalis??" ~ "Procambarus fallax f. virginalis",
      Species == "Trachemys scripta??" ~ "Trachemys scripta",
      Species == "Tradescantia fluminensis??" ~ "Tradescantia fluminensis",
      TRUE ~ Species
    )
  )


#defining colors for displaying Plants vs Animals
colors <- c(
  "Plant" = "#1b9e77",
  "Animal" = "#377eb8"
)

#adjusting the data to explore the development of species presence (yes/no) in trade
plot_species <- bind_rows(    #combining data
  unique.records.2021 %>%
    distinct(Species, Platform, Kingdom, Country) %>%
    mutate(year = "2021"),
  unique.records.2024 %>%
    distinct(Species, Platform, Kingdom, Country) %>%
    mutate(year = "2024")
)

species.summary <- plot_species %>%     #summarizing data
  count(Country, year, Platform, Kingdom)
species.summary <- species.summary %>%
  mutate(
    Kingdom = factor(Kingdom, levels = c("Plant", "Animal"))
  )

#malternatively this can be explored for unique records
plot_records <- bind_rows(
  unique.records.2021 %>%
    select(Species, Platform, Kingdom, Country) %>%
    mutate(year = "2021"),
  unique.records.2024 %>%
    select(Species, Platform, Kingdom, Country) %>%
    mutate(year = "2024")
)

records.summary <- plot_records %>%     #summarizing data
  count(Country, year, Platform, Kingdom)
records.summary <- species.summary %>%
  mutate(
    Kingdom = factor(Kingdom, levels = c("Plant", "Animal"))
  )

```

### Creating the underlying map

```{r}

europe <- ne_countries(scale = "medium", continent = "Europe", returnclass = "sf")

unique(europe$name)


# Crop to mainland Europe
europe_cropped <- st_crop(
  europe,
  xmin = -10, xmax = 35,
  ymin = 34, ymax = 72
)


Countries <- c("Germany", "Spain", "Czechia", "France", "Italy", "Netherlands")

map_data <- europe_cropped %>%
  filter(name %in% Countries)

centroids <- st_centroid(map_data)

centroids_df <- as.data.frame(st_coordinates(centroids))
centroids_df$Country <- map_data$name

#since we have "International data, we will add a manual location for it on the map
centroids_df <- rbind(
  centroids_df,
  data.frame(
    X = 25,   # longitude (adjust visually)
    Y = 45,   # latitude (adjust visually)
    Country = "International"
  )
)


```

### Creating functions

i\) a function to create individual stacked bar plots per country. The plots indicate the composition of animal and plant species found in Online retailers (left bar) and Peer-to-peer trade platforms (right) in 2021 and 2024

ii\) a function to display the individual bar plots on the underlying map

```{r}
make_barplot <- function(df, show_legend = FALSE) {
   df <- df %>%
    tidyr::complete(
      Platform,
      year,
      Kingdom = c("Plant", "Animal"),
      fill = list(n = 0)
    )
  ggplot(df, aes(x = Platform, y = n, fill = Kingdom)) +
    geom_col(position = "fill") +
    facet_wrap(~year) +
    scale_fill_manual(values = colors, drop = FALSE) +
     labs(
      x = "Platform",
      fill = "Kingdom"
    ) +
    theme_void()+
    theme(
      legend.position = if (show_legend) "bottom" else "none"
    )
}



make_map_plot <- function(summary_data, title_text) {

  base_map <- ggplot() +
    geom_sf(data = map_data, fill = "grey90", color = "black") +
    theme_void() +
    labs(title = title_text)

  for (i in 1:nrow(centroids_df)) {

    country_name <- centroids_df$Country[i]

    p <- make_barplot(filter(summary_data, Country == country_name))

    # Default offsets
x_offset <- -2.5
x_offset_max <- 2.5
y_offset <- -2.5
y_offset_max <- 2.5

# Netherlands: move left
if (country_name == "Netherlands") {
  x_offset <- -6
  x_offset_max <- -1
  y_offset <- -2
  y_offset_max <- 3
}

# Czechia: move right and up
if (country_name == "Czechia") {
  x_offset <- 1
  x_offset_max <- 6
  y_offset <- 0
  y_offset_max <- 5
}

base_map <- base_map +
  annotation_custom(
    grob = ggplotGrob(p),
    xmin = centroids_df$X[i] + x_offset,
    xmax = centroids_df$X[i] + x_offset_max,
    ymin = centroids_df$Y[i] + y_offset,
    ymax = centroids_df$Y[i] + y_offset_max
  )
  }
base_map <- base_map +
  annotate(
    "text",
    x = 25,
    y = 42,
    label = "International   kf",   #would not print the whole of International otherwise
    size = 3
  )
  return(base_map)
}
```

### Mapping the results

```{r}
map_species <- make_map_plot(
  species.summary,
  "Species Composition in e-commerce"
)

#creating a legend
legend_source <- make_barplot(
  species.summary %>% filter(Country == "Germany"),  # small subset is enough
  show_legend = TRUE
)
legend_only <- get_legend(legend_source)

#adding the legend to the figure
map_bar_species <- plot_grid(
  map_species,
  legend_only,
  ncol = 1,
  rel_heights = c(1, 0.1)
)
```

### Creating a heat map of change

The heat map will display the changes in plant and animal species composition per platform type and country. Specifically, the heat map shows the changes in plant proportions with positive values indicating more plants and negative values indicating more animals.

```{r}
make_heatmap <- function(summary_data, title_text) {

  heat_data <- summary_data %>%
    group_by(Country, year, Platform) %>%
    mutate(prop = n / sum(n)) %>%
    ungroup() %>%
    filter(Kingdom == "Plant") %>%   # adjust if your levels are "Plants"
    select(Country, Platform, year, prop) %>%
    tidyr::pivot_wider(
      names_from = year,
      values_from = prop
    ) %>%
    mutate(change = `2024` - `2021`)

  ggplot(heat_data, aes(x = Platform, y = Country, fill = change)) +
    geom_tile(color = "white") +
    scale_fill_gradient2(
      low = "#377eb8",   # more animals
      mid = "white",
      high = "#1b9e77",  # more plants
      midpoint = 0
    ) +
    labs(
      title = title_text,
      x = "Platform",
      y = "Country",
      fill = "Δ Plant proportion"
    ) +
    theme_minimal()+
  theme(legend.position = "right")
}

heat_species <- make_heatmap(
  species.summary,
  "Change in Plant Proportion (Unique Species) 2021-2024"
)


```
